---
title: "PDB Data Exploration"
author: "Alicia Key"
date: "`r Sys.Date()`"
output: html_document
---

# PDB Data Exploration

The Protein DataBank (PDB) stores files that contain the structure of "proteins, nucleic acids, and complex assemblies." These structures are essential tools for research in structural biology, biochemistry, and related fields. I was recently browsing Kaggle for datasets and found a scrape of PDB data up to the year 2018 by Shahir. The data included sequence information as well as metadata about those sequences. Kaggle suggests these data as a multi-class classification exercise. For this post, I am interested in exploring the metadata to find the data stored in the database and trends in how researchers collected those data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Include libraries and set plot theme.

### Libraries

```{r error=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(ggplot2))
```

### Set theme and `dplyr` options

```{r}
theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
```

## Load the dataset

```{r error=FALSE, warning=FALSE}
pdb <- read_csv("data/pdb_data_no_dups.csv.gz", show_col_types = FALSE)
```

### Look at the head of the dataset

```{r}
knitr::kable(head(pdb))
```

## Missing data

When I looked at the .csv in Excel, I found lots of missing data in the dataset. In this section, I quantify how much missing data there is.

### Bar graph of missing value counts

```{r}
pdb_total_rows = nrow(pdb)

pdb_missing <- pdb %>%
  summarize(
    structureIdNa = sum(is.na(structureId)) / pdb_total_rows * 100,
    classificationNa = sum(is.na(classification)) / pdb_total_rows * 100,
    experimentalTechniqueNa = sum(is.na(experimentalTechnique)) / pdb_total_rows * 100,
    macromoleculeTypeNa = sum(is.na(macromoleculeType)) / pdb_total_rows * 100,
    residueCountNa = sum(is.na(residueCount)) / pdb_total_rows * 100,
    resolutionNa = sum(is.na(resolution)) / pdb_total_rows * 100,
    structureMolecularWeightNa = sum(is.na(structureMolecularWeight)) / pdb_total_rows * 100,
    crystallizationMethodNa = sum(is.na(crystallizationMethod)) / pdb_total_rows * 100,
    crystallizationTempK = sum(is.na(crystallizationTempK)) / pdb_total_rows * 100,
    densityMatthewsNa = sum(is.na(densityMatthews)) / pdb_total_rows * 100,
    densityPercentSolNa = sum(is.na(densityPercentSol)) / pdb_total_rows * 100,
    pdbxDetailsNa = sum(is.na(pdbxDetails)) / pdb_total_rows * 100,
    phValueNa = sum(is.na(phValue)) / pdb_total_rows * 100,
    publicationYearNa = sum(is.na(publicationYear)) / pdb_total_rows * 100
  ) %>%
  pivot_longer(everything(), names_to = "column", values_to = "percent_missing") %>%
  arrange(percent_missing)

pdb_missing %>%
  mutate(
    column_sort = factor(
      column,
      levels = pdb_missing$column
    )
  ) %>%
  ggplot(aes(x = column_sort, y = percent_missing)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Percent missing per column",
      subtitle = "Higher is worse"
    )
```

### What if I dropped all `NA`?

Percentage observations remaining if all rows that have `NA` were dropped:

```{r}
pdb_no_na_count <- pdb %>%
  drop_na() %>%
  nrow()

pdb_no_na_count / pdb_total_rows * 100
```

Number of observations remaining if all the `NA`s were dropped.

```{r}
pdb_total_rows - pdb_no_na_count
```

### Plan of action: drop rows with `NA` and filter year range

For this exploration, I will simply drop all rows that have any `NA` variables. I will also restrict the year range to  2000 to 2018.

```{r}
pdb_clean <- pdb %>%
  drop_na() %>%
  filter(publicationYear >= 2000, publicationYear < 2018)
```

## Exploratory Visualization and Analysis

### Publications per macromolecule type, total up to 2018

Only drop the rows that have no `macromolecule` or `publicationYear` type.

Overall, it looks like proteins are by far the most dominant molecules being analyzed.

```{r}
pdb_publication_count_all_time <- pdb_clean %>%
  group_by(macromoleculeType) %>%
  summarize(publicationCount = n()) %>%
  arrange(desc(publicationCount))

knitr::kable(pdb_publication_count_all_time)
```

```{r}
ggplot(pdb_publication_count_all_time, aes(x = 1, y = publicationCount, fill = macromoleculeType)) +
  geom_col(position = "stack") +
  scale_fill_viridis_d() +
  labs(title = "Number of publications by macromolecule type") +
  ylab("publication count")
```

#### Publications per year, 2000 to 2017

```{r}
pdb_clean %>%
  group_by(publicationYear) %>%
  summarize(publicationCount = n()) %>%
  ggplot(aes(x = publicationYear, y = publicationCount)) +
    geom_line() +
    labs(title = "Publications per year") +
    ylab("publication count")
```

### Experimental techniques

#### Distribution of experimental techniques

Virtually all the publications in the cleaned dataset were from x-ray diffraction.

```{r}
pdb_techique_counts <- pdb_clean %>%
  group_by(experimentalTechnique) %>%
  summarize(publicationCount = n()) %>%
  arrange(desc(publicationCount))

knitr::kable(pdb_techique_counts)
```

#### Molecular weight by experimental technique

It looks like x-ray diffraction has the widest range of molecular weights...but there are far more x-ray diffraction observations in the cleaned dataset.

```{r}
ggplot(pdb_clean, aes(x = experimentalTechnique, y = structureMolecularWeight)) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip() +
  labs(
    title = "Molecular weight by experimental technique"
  ) +
  ylab("log(molecular weight)") +
  xlab("experimental technique")
```

### Other variables

#### Resolution by experimental technique

```{r}
ggplot(pdb_clean, aes(x = experimentalTechnique, y = resolution)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Resolution by experimental technique"
  ) +
  ylab("resolution (angstroms)") +
  xlab("experimental technique")
```

#### densityMatthews

Mathews density as described in [Matthews coefficient probabilities: Improved estimates for unit cell contents of proteins, DNA, and protein–nucleic acid complex crystals](https://onlinelibrary.wiley.com/doi/full/10.1110/ps.0350503).

```{r}
ggplot(pdb_clean, aes(x = densityMatthews)) +
  geom_histogram(bins = 100)
```

#### densityPercentSol

I wonder if these could be tied back to crystallization methods or crystallization temperature.

```{r}
ggplot(pdb_clean, aes(x = densityPercentSol)) +
  geom_histogram(bins = 100)
```

## Classification column

For this section, only apply cleaning to the `classification` column.

### Top classifications

What are the top 10 classifications?

```{r}
top_classifications <- pdb_clean %>%
  group_by(classification) %>%
  summarize(classification_count = n()) %>%
  arrange(desc(classification_count))

knitr::kable(head(top_classifications, 5))
```

## Conclusion

In the future, it would be interesting to:

- Train a model on the sequence data that accompanies this dataset, to see if I could fill in missing values based on sequence data.

- Find if there were any systematic features missing for each `macromoleculeType`. Maybe each macro molecule should get its own schema of attributes that are likely to be filled in.

- Mine the text of `pdbxDetails`. Perhaps these could correlated with things like ph, resolution, experimental method, etc.

- Mine the text of `crystallizationMethod`. Perhaps these could correlated with things like ph, resolution, experimental method, etc.

- Mine the text of `classification` since there are so many unique values in that column.

